---
title: "Section 2.1 Results"
output: html_notebook
---

# Introduction

This note book generates the figures for section 3.1 of the manuscript. 

### Load libraries & functions

```{r include=FALSE}
package_list <- c(
  "ShortRead", "tidyverse", "ggsci", "ggpubr",
  "cowplot", "patchwork", "readr", "writexl", "viridis", "patchwork"
)

lapply(package_list, require, character.only = TRUE)

results_dir <- "/mnt/HDD2/colin/ribosome_footprint_profiling/results/section2.1/"
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}
```

# Impact of TS on cell density
```{r}
cell_density_data <- read_delim("/mnt/HDD2/colin/ribosome_footprint_profiling/data/cell_density_data.txt",
  "\t",
  escape_double = FALSE, trim_ws = TRUE
)

cd_summary <- cell_density_data %>%
  group_by(Experiment, condition) %>%
  summarise(mean_density = mean(cell_density)) %>%
  group_by(Experiment)

cd_summary %>%
  filter(Experiment == "Elongation") %>%
  summarise(percent_cd_diff = 100 - (mean_density[condition == "TS"] / mean_density[condition == "NTS"]) * 100)

cd_summary %>%
  filter(Experiment == "Initiation") %>%
  summarise(percent_cd_diff = 100 - (mean_density[condition == "TS"] / mean_density[condition == "NTS"]) * 100)
```

## Cell density plot
```{r}
cell_density_plot <- ggboxplot(cell_density_data, x = "condition", y = "cell_density", color = "black", add = "jitter", fill = "condition") +
  facet_wrap(~Experiment, ncol = 2) +
  theme_bw() +
  stat_compare_means(method = "t.test", label.y = 2250000) +
  labs(x = "", y = "Cells/ml", fill = "") +
  theme(legend.position = "none") +
  scale_y_continuous(labels = function(x) format(x, scientific = TRUE)) +
  scale_fill_manual(values = c("#D55E00", "#56B4E9"))
ggsave(plot = cell_density_plot, filename = paste(results_dir, "Figure_S1.png", sep = ""), width = 6, height = 4, device = "png", dpi = 700)
cell_density_plot
```


# Analysis of read preprocessing 
## Ribo-seq
```{r}
samples <- c(
  "nts_r1", "nts_r2", "nts_r3", "nts_r4",
  "ts_r1", "ts_r2", "ts_r3", "ts_r4"
)

count_path <- "/mnt/HDD2/colin/ribosome_footprint_profiling/results/read_counts/"

positions <- c(
  "Cutadapt", "rRNA", "snoRNA", "tRNA", "<28nt or >32nt", "Retained"
)
```

### Cycloheximide
```{r include=FALSE}
riboseq_chx_raw <- read_delim(paste0(count_path, "riboseq_chx.raw.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Cycloheximide", stage = "raw")

riboseq_chx_trimmed <- read_delim(paste0(count_path, "riboseq_chx.trimmed.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Cycloheximide", stage = "trimmed")

riboseq_chx_rRNA <- read_delim(paste0(count_path, "riboseq_chx.rRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Cycloheximide", stage = "rRNA_retained")

riboseq_chx_tRNA <- read_delim(paste0(count_path, "riboseq_chx.tRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Cycloheximide", stage = "tRNA_retained")

riboseq_chx_snoRNA <- read_delim(paste0(count_path, "riboseq_chx.snoRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Cycloheximide", stage = "snoRNA_retained")

riboseq_chx_passed <- read_delim(paste0(count_path, "riboseq_chx.final.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Cycloheximide", stage = "pass")

chx_counts <- bind_rows(
  riboseq_chx_raw,
  riboseq_chx_trimmed,
  riboseq_chx_rRNA,
  riboseq_chx_tRNA,
  riboseq_chx_snoRNA,
  riboseq_chx_passed
) %>%
  select(
    sample_name,
    ribotype, stage,
    raw_read_number
  ) %>%
  pivot_wider(id_cols = c(sample_name, ribotype), names_from = stage, values_from = raw_read_number)

cycloheximide <- chx_counts %>%
  mutate(
    Cutadapt = raw - trimmed,
    rRNA = trimmed - rRNA_retained,
    snoRNA = rRNA_retained - snoRNA_retained,
    tRNA = snoRNA_retained - tRNA_retained,
    `<28nt or >32nt` = tRNA_retained - pass,
    Retained = pass
  ) %>%
  select(sample_name, ribotype, c("Cutadapt", "rRNA", "snoRNA", "tRNA", "<28nt or >32nt", "Retained"))
```

### Harringtonine
```{r include=FALSE}
riboseq_harr_raw <- read_delim(paste0(count_path, "riboseq_harr.raw.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Harringtonine", stage = "raw")

riboseq_harr_trimmed <- read_delim(paste0(count_path, "riboseq_harr.trimmed.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Harringtonine", stage = "trimmed")

riboseq_harr_rRNA <- read_delim(paste0(count_path, "riboseq_harr.rRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Harringtonine", stage = "rRNA_retained")

riboseq_harr_tRNA <- read_delim(paste0(count_path, "riboseq_harr.tRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Harringtonine", stage = "tRNA_retained")

riboseq_harr_snoRNA <- read_delim(paste0(count_path, "riboseq_harr.snoRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Harringtonine", stage = "snoRNA_retained")

riboseq_harr_passed <- read_delim(paste0(count_path, "riboseq_harr.final.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "Harringtonine", stage = "pass")

harr_counts <- bind_rows(
  riboseq_harr_raw,
  riboseq_harr_trimmed,
  riboseq_harr_rRNA,
  riboseq_harr_tRNA,
  riboseq_harr_snoRNA,
  riboseq_harr_passed
) %>%
  select(
    sample_name,
    ribotype, stage,
    raw_read_number
  ) %>%
  pivot_wider(id_cols = c(sample_name, ribotype), names_from = stage, values_from = raw_read_number)

harringtonine <- harr_counts %>%
  mutate(
    Cutadapt = raw - trimmed,
    rRNA = trimmed - rRNA_retained,
    snoRNA = rRNA_retained - snoRNA_retained,
    tRNA = snoRNA_retained - tRNA_retained,
    `<28nt or >32nt` = tRNA_retained - pass,
    Retained = pass
  ) %>%
  select(sample_name, ribotype, c("Cutadapt", "rRNA", "snoRNA", "tRNA", "<28nt or >32nt", "Retained"))
```
### No drug

```{r include=FALSE}
riboseq_nd_raw <- read_delim(paste0(count_path, "riboseq_nd.raw.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "No drug", stage = "raw")

riboseq_nd_trimmed <- read_delim(paste0(count_path, "riboseq_nd.trimmed.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "No drug", stage = "trimmed")

riboseq_nd_rRNA <- read_delim(paste0(count_path, "riboseq_nd.rRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "No drug", stage = "rRNA_retained")

riboseq_nd_tRNA <- read_delim(paste0(count_path, "riboseq_nd.tRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "No drug", stage = "tRNA_retained")

riboseq_nd_snoRNA <- read_delim(paste0(count_path, "riboseq_nd.snoRNA_filter.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "No drug", stage = "snoRNA_retained")

riboseq_nd_passed <- read_delim(paste0(count_path, "riboseq_nd.final.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "No drug", stage = "pass")

nd_counts <- bind_rows(
  riboseq_nd_raw,
  riboseq_nd_trimmed,
  riboseq_nd_rRNA,
  riboseq_nd_tRNA,
  riboseq_nd_snoRNA,
  riboseq_nd_passed
) %>%
  select(
    sample_name,
    ribotype, stage,
    raw_read_number
  ) %>%
  pivot_wider(id_cols = c(sample_name, ribotype), names_from = stage, values_from = raw_read_number)



nd <- nd_counts %>%
  mutate(
    Cutadapt = raw - trimmed,
    rRNA = trimmed - rRNA_retained,
    snoRNA = rRNA_retained - snoRNA_retained,
    tRNA = snoRNA_retained - tRNA_retained,
    `<28nt or >32nt` = tRNA_retained - pass,
    Retained = pass
  ) %>%
  select(sample_name, ribotype, c("Cutadapt", "rRNA", "snoRNA", "tRNA", "<28nt or >32nt", "Retained"))
```

```{r}
fn <- paste(results_dir, "Table S2.xlsx",
  sep = ""
)

suppressMessages(if (file.exists(fn)) {
  file.remove(fn)
})

write_xlsx(list(Cycloheximide = cycloheximide, Harrintonine = harringtonine, `No Drug` = nd),
  path = fn,
  format_headers = TRUE
)

bind_rows(cycloheximide, harringtonine, nd) %>%
  mutate(sample_name = case_when(
    sample_name == "nts_r1" ~ "NTS R1",
    sample_name == "nts_r2" ~ "NTS R2",
    sample_name == "nts_r3" ~ "NTS R3",
    sample_name == "nts_r4" ~ "NTS R4",
    sample_name == "ts_r1" ~ "TS R1",
    sample_name == "ts_r2" ~ "TS R2",
    sample_name == "ts_r3" ~ "TS R3",
    sample_name == "ts_r4" ~ "TS R4",
  )) %>%
  pivot_longer(cols = c("Cutadapt", "rRNA", "snoRNA", "tRNA", "<28nt or >32nt", "Retained"), values_to = "count", names_to = "stage") %>%
  ggplot(aes(x = sample_name, y = count, fill = factor(stage, levels = positions))) +
  geom_bar(stat = "identity", position = "fill") +
  theme_bw() +
  scale_fill_brewer(palette = "Paired") +
  labs(x = "", y = "Proportion of reads", fill = "") +
  facet_wrap(~ribotype, nrow = 3) +
  scale_fill_viridis(discrete = T)

ggsave(filename = paste(results_dir, "Figure_S2.png", sep = ""), width = 7, height = 8, device = "png", dpi = 700)
```
### Summary of reads per dataset
```{r}
bind_rows(chx_counts, harr_counts, nd_counts) %>%
  group_by(ribotype) %>%
  summarise(mean(raw))

bind_rows(chx_counts, harr_counts, nd_counts) %>%
  group_by(ribotype) %>%
  summarise(mean(1 - ((raw - tRNA_retained) / raw)))


bind_rows(chx_counts, harr_counts, nd_counts) %>%
  group_by(ribotype) %>%
  summarise(sum(pass))
```
## RNASeq
```{r}
rnaseq_raw <- read_delim(paste0(count_path, "rnaseq.raw.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "rnaseq", stage = "raw")


rnaseq_preprocessed <- read_delim(paste0(count_path, "rnaseq.preprocessed.counts"),
  "|",
  escape_double = FALSE, trim_ws = TRUE
) %>%
  mutate(sample_name = samples, ribotype = "rnaseq", stage = "pass")

rnaseq_counts <- bind_rows(
  rnaseq_raw,
  rnaseq_preprocessed
) %>%
  select(
    sample_name,
    ribotype,
    stage,
    raw_read_number
  ) %>%
  pivot_wider(id_cols = c(sample_name, ribotype), names_from = stage, values_from = raw_read_number)

rnaseq <- rnaseq_counts %>%
  mutate(
    Cutadapt = raw - pass,
    Retained = pass
  ) %>%
  select(sample_name, ribotype, c("Cutadapt", "Retained"))

write_xlsx(list(`RNA-seq` = rnaseq),
  path = fn,
  format_headers = TRUE
)



rnaseq_counts %>%
  summarise(mean(raw))

rnaseq %>%
  summarise(mean(Cutadapt), mean(Retained))
```
# RPF selected reads
```{r include=FALSE}
length_distribution_files <- fs::dir_ls("../read_length_distribution/", 
                        regexp = "\\.txt$",)

length_distirbution_data <- length_distribution_files %>%
  map_dfr(read_delim, " ", escape_double=FALSE,trim_ws=TRUE, .id = "source", col_names = FALSE) %>%
  mutate(ribotype=case_when(
    str_detect(source,"riboseq_chx") ~ "riboseq_chx",
    str_detect(source,"riboseq_harr") ~ "riboseq_harr",
    str_detect(source,"riboseq_nd") ~ "riboseq_nd",
    str_detect(source,"rnaseq_se") ~ "RNA-seq")) %>%
  mutate(sample=case_when(
    str_detect(source,"nts_r1") ~ "NTS_R1",
     str_detect(source,"nts_r2") ~ "NTS_R2",
     str_detect(source,"nts_r3") ~ "NTS_R3",
     str_detect(source,"nts_r4") ~ "NTS_R4",
       str_detect(source,"ts_r1") ~ "TS_R1",
     str_detect(source,"ts_r2") ~ "TS_R2",
     str_detect(source,"ts_r3") ~ "TS_R3",
     str_detect(source,"ts_r4") ~ "TS_R4",
    str_detect(source,"merged") ~ "rnaseq")) %>%
  mutate(readlength=X2,
         count=X1) %>%
  dplyr::select(sample,ribotype,readlength, count)

table(length_distirbution_data$ribotype)
```

```{r}
labels <- c(riboseq_chx = "Ribo-seq (CHX)", riboseq_harr = "Ribo-seq (HARR)", riboseq_nd = "Ribo-seq (ND)", `RNA-seq` = "RNA-seq")

read_lengths_hist_plot <- length_distirbution_data %>%
  filter(readlength >= 25 & readlength <= 34) %>%
  mutate(ribotype = factor(ribotype, levels = c("riboseq_harr", "riboseq_chx", "riboseq_nd", "RNA-seq"))) %>%
  group_by(ribotype) %>%
  mutate(total = count) %>%
  ggplot(aes(x = readlength, y = total, fill = ifelse(readlength > 27 & readlength < 32 & ribotype != "RNA-seq", "RPF", "RNA"))) +
  geom_bar(stat = "identity") +
  facet_wrap(~ribotype, ncol = 2, labeller = labeller(ribotype = labels), scales = "free_x") +
  labs(y = "Number of reads", x = "Read length (nts)", fill = "") +
  scale_y_continuous(breaks = c(20000000, 40000000, 60000000), labels = c("20 million", "40 million", "60 million")) +
  scale_x_continuous(breaks = 25:34) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"), legend.position = "bottom") +
  scale_fill_viridis(discrete = T, direction = -1)

read_lengths_hist_plot
ggsave(plot = read_lengths_hist_plot, filename = paste(results_dir, "Figure_3.1_C.png", sep = ""), width = 5, height = 5, device = "png", dpi = 700)
```
## Phasing
```{r include=FALSE}
phasing_path <- "/mnt/HDD2/colin/ribosome_footprint_profiling/rpf_psite_analysis/"

riboseq_chx_phasing <- read_delim(paste0(phasing_path, "riboseq_chx_phasing.txt"),
  "\t",
  escape_double = FALSE, trim_ws = TRUE,
  skip = 32
) %>%
  summarise(
    `0` = mean(phase0),
    `1` = mean(phase1),
    `2` = mean(phase2)
  ) %>%
  mutate(ribotype = "Ribo-seq (CHX)")

riboseq_harr_phasing <- read_delim(paste0(phasing_path, "riboseq_harr_phasing.txt"),
  "\t",
  escape_double = FALSE, trim_ws = TRUE,
  skip = 32,
) %>%
  summarise(
    `0` = mean(phase0),
    `1` = mean(phase1),
    `2` = mean(phase2)
  ) %>%
  mutate(ribotype = "Ribo-seq (HARR)")

riboseq_nd_phasing <- read_delim(paste0(phasing_path, "riboseq_nd_phasing.txt"),
  "\t",
  escape_double = FALSE, trim_ws = TRUE,
  skip = 32
) %>%
  summarise(
    `0` = mean(phase0),
    `1` = mean(phase1),
    `2` = mean(phase2)
  ) %>%
  mutate(ribotype = "Ribo-seq (ND)")

rnaseq_phasing <- read_delim(paste0(phasing_path, "rnaseq_se_phasing.txt"),
  "\t",
  escape_double = FALSE, trim_ws = TRUE,
  skip = 32
) %>%
  summarise(
    `0` = mean(phase0),
    `1` = mean(phase1),
    `2` = mean(phase2)
  ) %>%
  mutate(ribotype = "RNA-seq")
```

## Plot the phasing figure 
```{r}
phasing_bar_plot <- bind_rows(riboseq_chx_phasing, riboseq_harr_phasing, riboseq_nd_phasing, rnaseq_phasing) %>%
  pivot_longer(cols = c(`0`, `1`, `2`), values_to = "Proportion") %>%
  mutate(ribotype = factor(ribotype, levels = c("Ribo-seq (HARR)", "Ribo-seq (CHX)", "Ribo-seq (ND)", "RNA-seq"))) %>%
  ggplot(aes(x = name, y = Proportion, fill = name)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ribotype, ncol = 2) +
  labs(x = "Frame", y = "Proportion of RPFs", fill = "Frame") +
  scale_fill_viridis(discrete = T) +
  theme_bw() +
  theme(strip.background = element_rect(fill = "white"), legend.position = "bottom")

phasing_bar_plot

ggsave(plot = phasing_bar_plot, filename = paste(results_dir, "Figure_1E.png", sep = ""), width = 5, height = 5, device = "png", dpi = 700)
```
## Enrichment of RPFs in the harringtonine data at annotated TIS
Demonstration that harringtonine has a higher occupancy for annotation TIS in comparison to CHX and No-drug
```{r}
chx_metagene <- read_tsv("../../orfrater_analysis/chx/metagene.txt") %>%
  mutate(ribotype = "chx")
harr_metagene <- read_tsv("../../orfrater_analysis/harr/metagene.txt") %>%
  mutate(ribotype = "harr")
nd_metagene <- read_tsv("../../orfrater_analysis/nd/metagene.txt") %>%
  mutate(ribotype = "nd")
metagene <- bind_rows(chx_metagene, harr_metagene, nd_metagene)

metagene <- metagene %>%
  mutate(Means = rowMeans(.[, 3:6]))

metagene <- metagene %>%
  mutate(Frame = case_when(
    position %% 3 == 0 ~ "0",
    position %% 3 == 1 ~ "1",
    position %% 3 == 2 ~ "2"
  ))

metagene$region <- factor(metagene$region, # Reordering group factor levels
  levels = c("START", "CDS", "STOP")
)

labels <- c(chx = "Cycloheximide", harr = "Harringtonine", nd = "No drug")

safe_colorblind_palette <- c("#009E73", "#D55E00", "#999999")

start_codon_plot <- metagene %>%
  filter(region == "START") %>%
  filter(position <= 50) %>%
  mutate(ribotype = case_when(
    ribotype == "chx" ~ "Cycloheximide",
    ribotype == "harr" ~ "Harringtonine",
    ribotype == "nd" ~ "No drug"
  )) %>%
  mutate(ribotype = factor(ribotype, levels = c("Harringtonine", "Cycloheximide", "No drug"))) %>%
  ggplot(aes(x = position, y = Means, color = ribotype), alpha = ribotype) +
  geom_line(size = 0.4) +
  geom_point(size = 0.8) +
  labs(y = "Averge RPF density", color = "", x = "Transcript position") +
  theme_bw() +
  theme(legend.position = "top") +
  scale_color_viridis(discrete = T) +
  # scale_color_manual(values=safe_colorblind_palette,
  #                     limits = c("chx", "harr", "nd"),
  #   labels = c("Cycloheximide", "Harringtonine", "No drug")) +
  scale_alpha_manual(values = c(0.6, 1, 0.6))
start_codon_plot
ggsave(filename = paste(results_dir, "Figure_1F.png", sep = ""), width = 6, height = 4, device = "png", dpi = 700)
```

# Knit to R script
```{r}
#knitr::purl("Results_2_1.Rmd")
```
